### 2016-06-30 Projects/DelCo_2012/Report/DelCo_2012.tex
fix SS/SB wording in CAFS

\begin{tabbing}
\hspace{1.5in} \= \\
CAFS \> Compressed Air Foam System \\
ESTC \> Delaware County Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
TFT \> Task Force Tips \\
\end{tabbing}

@@

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAFS for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAFS for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAFS for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAFS and plain water to evaluate the effectiveness of CAFS for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAFS and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. The research findings are provide information with the potential to improve the safe and effective use of CAFS for structural firefighting.

@@

The key measures were to understand the cooling and suppression capabilities of CAFS compared to water. The experimental designs were developed based on a project planning workshop which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAFS during interior attack operations in residential occupancies~\cite{Grant:2011}. Anecdotally, other fire departments or firefighters had similar incidents or impressions of CAFS.

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allows the water to penetrate the hot surface of a burning fuel, and also provides additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAFS increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAFS advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

@@

There has never been a comprehensive scientific study on the use of CAFS for structural fire fighting.  Fire departments have conducted pilot programs to examine implementation and effectiveness of CAFS, foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures, and test laboratories and universities have examined small sections of the CAFS puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}.

@@

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland firefighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural firefighting, crews typically  have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they may have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior firefighting with CAFS are understood better.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in the State of Maryland and was a committed user of CAFS with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires whiles CAFS was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAFS to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of (MCFRS) issued a General Order prohibiting the use of CAFS during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAFS during structure fires and to report back his findings and recommendations. The catalyst for this study was as a result of this order.

@@

All of the experiments were conducted at full scale in compartments and structures that were representitive room sizes in a residential occupancy.  Technical panel members whose departments rountinely use CAFS for structural firefighting were invited and present at many of the fire suppression experiments.  The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012.  The data from these experiments were presented to the technical panel member.   The fire suppression experiments were conducted with a fuel package that included upholstered furniture.  After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package was composed mainly of wood based materials, as a way to increase the challenge of the suppression experiment.  After the wood based fuel package experiments were completed, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline.  The issue of a larger fire compartment had also been raised durning the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan.  In order to accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015.

The sections below describe the types of experiments, the experimental facilities, instrumentation, and fuel loads.  Project partner, Montgomery County (MD) Fire and Rescue Service (MCFRS) provided the CAFS equipped engines and the staffing needed to provide both the water and CAFS for the experiments. These experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.   Baseline operating conditions were agreed upon at the workshop for these experiments (Table~\ref{tab:op_condition}).   Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}.   The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120 gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior.

@@

The fire suppression series included 18 tests in two single-story experimental structures and one two-story structure. Fourteen tests were conducted in the single story structures (7 tests in each structure) using both water and CAFS application as a suppression agent. The nozzle patterns were varied from straight stream to narrow fog. Two of the tests examined suppression tactics for attic fires in residential dwellings, one in each structure. For the attic tests, a smooth bore nozzle was used. The 4 two-story tests used both water and CAFS from a straight-stream nozzle.

@@

Spray density experiments were conducted by placing 0.23~m$^2$ (2.5~ft$^2$) interlocking collector bins across the floor surface area of the burn rooms within each structure. The buildings were sheathed in plastic for the spray density tests to protect and preserve as much of the structure as possible before any fire suppression tests were conducted. Water flowed for a specified duration and upon completion the bins were removed from the structure. Each bin was then weighed to quantify the impact of nozzle location and spray pattern. Thirty-three spray density tests were conducted; 21 were in the burn building for the gas cooling tests and 12 were in the concrete structure for the suppression experiments. The flow rate for each test was 120~gpm for water and 120~gpm plus 60~cfm for CAFS. Tables\ref{tab:spray_density_tests} and \ref{tab:spray_density_tests2} provide an overview of each experiment. In the table, GC stands for the gas cooling experiments while SE stands for the suppression experiments. For all of the experiments, 100~ft (30~m) of 1~3/4~inch hoseline connected to a Blitz Fire monitor with a Metro 1 nozzle was used.

@@

\begin{table}[!ht]
\footnotesize
\centering
\captionof{table}{Spray Density Test Matrix for Gas Cooling Experiments}\label{tab:spray_density_tests}
\begin{tabular}{lllll}
\toprule[1.5pt]
Test \#    &  Duration (s)  & Agent  &  Pattern            & Nozzle Position  \\
\midrule
GC1        &  15            & Water  &  Solid Stream       & Back             \\
GC2        &  23            & Water  &  Solid Stream       & Back             \\
GC3        &  15            & Water  &  Solid Stream       & Mid              \\
GC4        &  15            & Water  &  Solid Stream       & Mid              \\
GC5        &  20            & Water  &  Solid Stream       & Mid              \\
GC6        &  21            & Water  &  30$^{\circ}$ Fog   & Mid              \\
GC7        &  20            & Water  &  30$^{\circ}$ Fog   & Mid              \\
GC8        &  22            & Water  &  30$^{\circ}$ Fog   & Back             \\
GC9        &  24            & Water  &  30$^{\circ}$ Fog   & Back             \\
GC10       &  20            & Water  &  Solid 7/8~in Slug  & Back             \\
GC11       &  22            & Water  &  Solid 7/8~in Slug  & Back             \\
GC12       &  15            & CAFS    &  Smooth Bore 7/8~in & Back             \\
GC13       &  10            & CAFS    &  Smooth Bore 7/8~in & Mid              \\
GC14       &  15            & CAFS    &  Smooth Bore 7/8~in & Back             \\
GC15       &  20            & CAFS    &  Smooth Bore 7/8~in & Mid              \\
GC16       &  15            & CAFS    &  30$^{\circ}$ Fog   & Mid              \\
GC17       &  20            & CAFS    &  Solid Stream       & Mid             \\
GC18       &  10            & CAFS    &  Solid Stream       & Mid             \\
GC19       &  15            & CAFS    &  30$^{\circ}$ Fog   & Back             \\
GC20       &  15            & CAFS    &  30$^{\circ}$ Fog   & Back             \\
GC21       &  15            & CAFS    &  Solid Stream       & Back             \\

@@

\begin{table}[!ht]
\footnotesize
\centering
\captionof{table}{Spray Density Test Matrix for Fire Suppression Experiments}\label{tab:spray_density_tests2}
\begin{tabular}{lllll}
\toprule[1.5pt]
Test \#    & Duration (s)  & Agent  &  Pattern            & Nozzle Location  \\
\midrule
SE1        & 7             & Water  &  Solid Stream       &    Fire Room          \\
SE2        & 18            & Water  &  Solid Stream       &    Fire Room          \\
SE3        & 19            & Water  &  Solid Stream       &    Fire Room          \\
SE4        & 21            & Water  &  Solid Stream       &    Fire Room          \\
SE5        & 22            & Water  &  30$^{\circ}$ Fog   &    Fire Room          \\
SE6        & 19            & Water  &  30$^{\circ}$ Fog   &    Fire Room          \\
SE7        & 16            & Water  &  Solid 7/8~in Slug  &    Fire Room          \\
SE8        & 17            & Water  &  Solid 7/8~in Slug  &    Hallway            \\
SE9        & 15            & Water  &  30$^{\circ}$ Fog   &    Hallway            \\
SE10       & 16            & Water  &  30$^{\circ}$ Fog   &    Hallway            \\
SE11       & 16            & Water  &  Solid Stream       &    Hallway            \\
SE12       & 16            & Water  &  Solid Stream       &    Hallway            \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

The 88 gas cooling events were completed as part of 9 separate tests in the burn building. During each test, the fire was ignited and then developed until the temperature 1.83~m (6~ft) below the ceiling in the gas cooling room reached or exceeded 250~$^{\circ}$C (482~$^{\circ}$F). Once this criteria was met, gas cooling began. Figure~\ref{fig:gas_cooling_exp4} shows 19 cooling events at one thermocouple array during a continuous test.

@@

There were three thermocouple arrays in the gas cooling room (adjacent to the burn room in Fig.~\ref{fig:Gas_Cooling_Instrumentation_Dimensions}). Figures~\ref{fig:gas_cooling_sub1} and \ref{fig:gas_cooling_sub5} show the variation of the impact of the suppression stream on cooling location for three consecutive water and CAFS tests, respectively. For both sets a solid 7/8 in nozzle in the middle position was used. The data shows that thermocouples in the plot in the upper right of both figures has more significant temperature drops than the other two arrays; this is likely a result of that thermocouple array being directly ``hit'' by the suppression stream.

@@

To better quantify the effectiveness of the suppression agents (CAFS, water), the maximum temperature differences associated with each gas cooling event for each agent from the three thermocouple trees were analyzed. Figure~\ref{fig:combined_all} compares the temperature differences at each thermocouple position (via a scatterplot) for the 30 replicate tests, 15 water and 15 CAFS, for all streams aimed at all positions.

@@

The circles in Fig.~\ref{fig:combined_all} are the maximum temperature differences for each measurement point for all the replicate cooling events. If water and CAFS have the exact same temperature differences, all of the points would fall along solid black line in the figure. Circles that vary from that line represent measurements where either water (above) or CAFS (below) cooled more significantly than the other agent. The gray area surrounding the solid black line, represents measurement uncertainty (\ref{subsec:Uncertainty}). For data that fall within this area, there is an inability to determine which agent was more or less effective than the other. The red dashed line is the line the describes the data by minimizing the differences between the two suppression agents through a linear approximation of the data. Since this line falls above the gray area for the comparison, water shows a better ability to cool gas under the same test criteria as CAFS.

@@

For the 12 middle position replicate comparisons (Fig.~\ref{fig:CAFS_Water_mid}), the results (via the dashed red line) show that water has a slight advantage in gas cooling relative to CAFS. At the back position (Fig.~\ref{fig:CAFS_Water_full}), for the 3 replicate comparison, the least squared error line falls within the experimental uncertainty. This indicates an inability to distinguish between the two suppression agents with statistical confidence.

To further analyze the data, the temperature differences can be separated by each of the three thermocouples in the fire room. Figure~\ref{fig:Gas_Cooling_Instrumentation_Dimensions2} shows the placement of three thermocouple arrays (TC A1, TC A2, and TC A3).
\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/PDFs/West_Structure/DelCo_2012_West_Structure_Instrumentation2}
	\caption{Schematic of  the Gas Cooling Experiments.}
	\label{fig:Gas_Cooling_Instrumentation_Dimensions2}
\end{figure}
Based on the monitor position (arrow at the doorway in Fig.~\ref{fig:Gas_Cooling_Instrumentation_Dimensions2}), thermocouple arrays A1 and A2 had more direct impact from the suppression agent compared to thermocouple array A3. Figure~\ref{fig:Burn_Building_Test_18b} shows the distribution of the agent within the burn building from a spray density test for a straight stream with CAFS at the mid position compared to the positions of the thermocouple trees.
\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Bars/BB18b}
	\caption{Gas Cooling Spray Density Test 18}
	\label{fig:Burn_Building_Test_18b}
\end{figure}

There is a difference in the amount of direct impact from the suppression agent at the three spatial locations. Therefore, the maximum temperature differences for each thermocouple on each of the three arrays should also be analyzed separately. Figures~\ref{fig:CAFS_Water_A1_all}, \ref{fig:CAFS_Water_A2_all}, and \ref{fig:CAFS_Water_A3_all} show the temperature cooling scatter for thermocouple trees A1, A2, and A3 respectively. Note that thermocouple array A3, the array which has the least amount of direct suppression, has the closest fit compared to the solid black line that represents uniform impact for both suppression agents. The fit lines for A1 and A3 also fall within the experimental uncertainty, meaning that differences in the impact of CAFS and water are indistinguishable. Thermocouple tree A2, which gets the most significant direct impact from the suppression agents, shows a bias towards water being more effective than CAFS.

@@

Note that while there are some differences in the changes in temperature and heat flux when comparing CAFS to Water, differences are not substantial. Both suppression agents had similar, significant impact in reducing the high-temperature, high-heat flux environment that existed before suppression. The first (hallway) suppression, an indirect attack, did a majority of the cooling in all of the tests. This illustrates the impact of gas cooling, independent of suppression agent. In several cases rekindling occurred, as seen by the temperature increase between the two nozzle applications, but all cases the second (room) attack controlled the fire.

@@

The gas cooling experiments were designed to measure the ability of water streams and CAFS streams to cool a hot gas layer with a minimum temperature of  250~$^{\circ}$C (482~$^{\circ}$F) at positions 1.83~m (6~ft) below the ceiling. During the planning of the experiments, some firefighters thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAFS, since the injected air, air being a great insulator, would interfere and result in less cooling. It was the belief that the reduced gas cooling capabilities of CAFS may have lead to firefighter injuries during interior fire attacks. Other firefighters thought that the bubble structure would provide the same mass flowrate of water with more surface area, which would enable a faster heat transfer rate and therefore, improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.

>>>>

\begin{tabbing}
\hspace{1.5in} \= \\
CAF \> Compressed Air Foam \\
CAFS \> Compressed Air Foam System \\
ESTC \> Delaware County Emergency Services Training Center \\
MCFRS \> Montgomery County (MD) Fire and Rescue Service \\
NIST \> National Institute of Standards and Technology \\
TFT \> Task Force Tips \\
\end{tabbing}

@@

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAFS for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAF for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAF for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. The research findings are provide information with the potential to improve the safe and effective use of CAF for structural firefighting.

@@

The key measures were to understand the cooling and suppression capabilities of CAF compared to water. The experimental designs were developed based on a project planning workshop which included the discussion of incidents where firefighters had concerns or experienced minor burns while using CAF during interior attack operations in residential occupancies~\cite{Grant:2011}. Anecdotally, other fire departments or firefighters had similar incidents or impressions of CAF.

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allows the water to penetrate the hot surface of a burning fuel, and also provides additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAF advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

@@

There has never been a comprehensive scientific study on the use of CAF for structural fire fighting. Fire departments have conducted pilot programs to examine implementation and effectiveness of CAF, foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures, and test laboratories and universities have examined small sections of the CAF puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}.

@@

In 2005, The National Fallen Firefighters Foundation held the National Fire Service Research Agenda Symposium.  One of the needs identified was ``improved extinguishing agents and delivery systems.'' The report states that ``The primary objective of this research should be to reduce the exposure of firefighters to the risks of interior fire suppression by developing more effective extinguishing agents and systems.'' In wildland firefighting, CAFS is typically applied directly to the fuel from a remote location, typically a non-IDLH environment. In structural firefighting, crews typically  have to enter the building, which contains the fire, and move under or through hot combustion products and pre-heated areas of the building (IDLH environment) as they approach the fire. In many cases, before the firefighters are able to apply the suppression agent directly to the burning fuel, they may have to mitigate the hazard of the hot gas layer or try to cool interior surfaces as they move toward the fire. It is important that these distinctions and differences between exterior and interior firefighting with CAF are understood better.

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in the State of Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires whiles CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of (MCFRS) issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. The catalyst for this study was as a result of this order.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments. The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel member. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package was composed mainly of wood based materials, as a way to increase the challenge of the suppression experiment. After the wood based fuel package experiments were completed, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan.  In order to accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015.

The sections below describe the types of experiments, the experimental facilities, instrumentation, and fuel loads. Project partner, Montgomery County (MD) Fire and Rescue Service (MCFRS) provided the CAFS equipped engines and the staffing needed to provide both the water and CAF for the experiments. These experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression. Baseline operating conditions were agreed upon at the workshop for these experiments (Table~\ref{tab:op_condition}). Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior.

@@

The fire suppression series included 18 tests in two single-story experimental structures and one two-story structure. Fourteen tests were conducted in the single story structures (7 tests in each structure) using both water and CAF application as a suppression agent. The nozzle patterns were varied from straight stream to narrow fog. Two of the tests examined suppression tactics for attic fires in residential dwellings, one in each structure. For the attic tests, a smooth bore nozzle was used. The 4 two-story tests used both water and CAF from a straight-stream nozzle.

@@

Spray density experiments were conducted by placing 0.23~m$^2$ (2.5~ft$^2$) interlocking collector bins across the floor surface area of the burn rooms within each structure. The buildings were sheathed in plastic for the spray density tests to protect and preserve as much of the structure as possible before any fire suppression tests were conducted. Water flowed for a specified duration and upon completion the bins were removed from the structure. Each bin was then weighed to quantify the impact of nozzle location and spray pattern. Thirty-three spray density tests were conducted; 21 were in the burn building for the gas cooling tests and 12 were in the concrete structure for the suppression experiments. The flow rate for each test was 120~gpm for water and 120~gpm plus 60~cfm for CAF. Tables\ref{tab:spray_density_tests} and \ref{tab:spray_density_tests2} provide an overview of each experiment. In the table, GC stands for the gas cooling experiments while SE stands for the suppression experiments. For all of the experiments, 100~ft (30~m) of 1~3/4~inch hoseline connected to a Blitz Fire monitor with a Metro 1 nozzle was used.

@@

\begin{table}[!ht]
\footnotesize
\centering
\captionof{table}{Spray Density Test Matrix for Gas Cooling Experiments}\label{tab:spray_density_tests}
\begin{tabular}{lllll}
\toprule[1.5pt]
Test \#    &  Duration (s)  & Agent  &  Pattern            & Nozzle Position  \\
\midrule
GC1        &  15            & Water  &  Straight Stream    & Back             \\
GC2        &  23            & Water  &  Straight Stream    & Back             \\
GC3        &  15            & Water  &  Straight Stream    & Mid              \\
GC4        &  15            & Water  &  Straight Stream    & Mid              \\
GC5        &  20            & Water  &  Straight Stream    & Mid              \\
GC6        &  21            & Water  &  30$^{\circ}$ Fog   & Mid              \\
GC7        &  20            & Water  &  30$^{\circ}$ Fog   & Mid              \\
GC8        &  22            & Water  &  30$^{\circ}$ Fog   & Back             \\
GC9        &  24            & Water  &  30$^{\circ}$ Fog   & Back             \\
GC10       &  20            & Water  &  Solid Stream       & Back             \\
GC11       &  22            & Water  &  Solid Stream       & Back             \\
GC12       &  15            & CAF    &  Solid Stream       & Back             \\
GC13       &  10            & CAF    &  Solid Stream       & Mid              \\
GC14       &  15            & CAF    &  Solid Stream       & Back             \\
GC15       &  20            & CAF    &  Solid Stream       & Mid              \\
GC16       &  15            & CAF    &  30$^{\circ}$ Fog   & Mid              \\
GC17       &  20            & CAF    &  Straight Stream    & Mid             \\
GC18       &  10            & CAF    &  Straight Stream    & Mid             \\
GC19       &  15            & CAF    &  30$^{\circ}$ Fog   & Back             \\
GC20       &  15            & CAF    &  30$^{\circ}$ Fog   & Back             \\
GC21       &  15            & CAF    &  Straight Stream    & Back             \\

@@

\begin{table}[!ht]
\footnotesize
\centering
\captionof{table}{Spray Density Test Matrix for Fire Suppression Experiments}\label{tab:spray_density_tests2}
\begin{tabular}{lllll}
\toprule[1.5pt]
Test \#    & Duration (s)  & Agent  &  Pattern            & Nozzle Location  \\
\midrule
SE1        & 7             & Water  &  Straight Stream    &    Fire Room          \\
SE2        & 18            & Water  &  Straight Stream    &    Fire Room          \\
SE3        & 19            & Water  &  Straight Stream    &    Fire Room          \\
SE4        & 21            & Water  &  Straight Stream    &    Fire Room          \\
SE5        & 22            & Water  &  30$^{\circ}$ Fog   &    Fire Room          \\
SE6        & 19            & Water  &  30$^{\circ}$ Fog   &    Fire Room          \\
SE7        & 16            & Water  &  Solid Stream       &    Fire Room          \\
SE8        & 17            & Water  &  Solid Stream       &    Hallway            \\
SE9        & 15            & Water  &  30$^{\circ}$ Fog   &    Hallway            \\
SE10       & 16            & Water  &  30$^{\circ}$ Fog   &    Hallway            \\
SE11       & 16            & Water  &  Straight Stream    &    Hallway            \\
SE12       & 16            & Water  &  Straight Stream    &    Hallway            \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

The 88 gas cooling events were completed as part of 9 separate tests in the burn building. For detailed discussion on each of the individual events, please refer to a prior analysis conducted by Mitchell \cite{Mitchell:1}. The focus of this analysis is to look at trends within the aggregate data. During each test, the fire was ignited and then developed until the temperature 1.83~m (6~ft) below the ceiling in the gas cooling room reached or exceeded 250~$^{\circ}$C (482~$^{\circ}$F). Once this criteria was met, gas cooling began. Figure~\ref{fig:gas_cooling_exp4} shows 19 cooling events at one thermocouple array during a continuous test.

@@

There were three thermocouple arrays in the gas cooling room (adjacent to the burn room in Fig.~\ref{fig:Gas_Cooling_Instrumentation_Dimensions}). Figures~\ref{fig:gas_cooling_sub1} and \ref{fig:gas_cooling_sub5} show the variation of the impact of the suppression stream on cooling location for three consecutive water and CAF tests, respectively. For both sets a solid 7/8~in nozzle in the middle position was used. The data shows that thermocouples in the plot in the upper right of both figures has more significant temperature drops than the other two arrays; this is likely a result of that thermocouple array being directly ``hit'' by the suppression stream.

@@

To better quantify the effectiveness of the suppression agents (CAF, water), the maximum temperature differences associated with each gas cooling event for each agent from the three thermocouple trees were analyzed. Figure~\ref{fig:combined_all} compares the temperature differences at each thermocouple position (via a scatterplot) for the 30 replicate tests, 15 water and 15 CAF, for all streams aimed at all positions.

@@

The circles in Fig.~\ref{fig:combined_all} are the maximum temperature differences for each measurement point for all the replicate cooling events. If water and CAFS have the exact same temperature differences, all of the points would fall along solid black line in the figure. Circles that vary from that line represent measurements where either water (above) or CAF (below) cooled more significantly than the other agent. The gray area surrounding the solid black line, represents measurement uncertainty (\ref{subsec:Uncertainty}). For data that fall within this area, there is an inability to determine which agent was more or less effective than the other. The red dashed line is the line the describes the data by minimizing the differences between the two suppression agents through a linear approximation of the data. Since this line falls above the gray area for the comparison, water shows a better ability to cool gas under the same test criteria as CAF.

@@

For the 12 middle position replicate comparisons (Fig.~\ref{fig:CAFS_Water_mid}), the results (via the dashed red line) show that water has a slight advantage in gas cooling relative to CAF. At the back position (Fig.~\ref{fig:CAFS_Water_full}), for the 3 replicate comparison, the least squared error line falls within the experimental uncertainty. This indicates an inability to distinguish between the two suppression agents with statistical confidence.

To further analyze the data, the temperature differences can be separated by each of the three thermocouples in the fire room. Figure~\ref{fig:Gas_Cooling_Instrumentation_Dimensions2} shows the placement of three thermocouple arrays (TC A1, TC A2, and TC A3).
\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/PDFs/West_Structure/DelCo_2012_West_Structure_Instrumentation2}
	\caption{Schematic of  the Gas Cooling Experiments.}
	\label{fig:Gas_Cooling_Instrumentation_Dimensions2}
\end{figure}
Based on the monitor position (arrow at the doorway in Fig.~\ref{fig:Gas_Cooling_Instrumentation_Dimensions2}), thermocouple arrays A1 and A2 had more direct impact from the suppression agent compared to thermocouple array A3. Figure~\ref{fig:Burn_Building_Test_18b} shows the distribution of the agent within the burn building from a spray density test for a straight stream with CAF at the mid position compared to the positions of the thermocouple trees.
\begin{figure}[!ht]
	\includegraphics[width=.8\columnwidth]{../Figures/Bars/BB18b}
	\caption{Gas Cooling Spray Density Test 18}
	\label{fig:Burn_Building_Test_18b}
\end{figure}

There is a difference in the amount of direct impact from the suppression agent at the three spatial locations. Therefore, the maximum temperature differences for each thermocouple on each of the three arrays should also be analyzed separately. Figures~\ref{fig:CAFS_Water_A1_all}, \ref{fig:CAFS_Water_A2_all}, and \ref{fig:CAFS_Water_A3_all} show the temperature cooling scatter for thermocouple trees A1, A2, and A3 respectively. Note that thermocouple array A3, the array which has the least amount of direct suppression, has the closest fit compared to the solid black line that represents uniform impact for both suppression agents. The fit lines for A1 and A3 also fall within the experimental uncertainty, meaning that differences in the impact of CAF and water are indistinguishable. Thermocouple tree A2, which gets the most significant direct impact from the suppression agents, shows a bias towards water being more effective than CAF.

@@

Note that while there are some differences in the changes in temperature and heat flux when comparing CAF to Water, differences are not substantial. Both suppression agents had similar, significant impact in reducing the high-temperature, high-heat flux environment that existed before suppression. The first (hallway) suppression, an indirect attack, did a majority of the cooling in all of the tests. This illustrates the impact of gas cooling, independent of suppression agent. In several cases rekindling occurred, as seen by the temperature increase between the two nozzle applications, but all cases the second (room) attack controlled the fire.

@@

The gas cooling experiments were designed to measure the ability of water streams and CAF streams to cool a hot gas layer with a minimum temperature of 250~$^{\circ}$C (482~$^{\circ}$F) at positions 1.83~m (6~ft) below the ceiling. During the planning of the experiments, some firefighters thought that the excellent heat absorbing capacity of plain water would exceed the heat absorbing capabilities of CAFS, since the injected air, air being a great insulator, would interfere and result in less cooling. It was the belief that the reduced gas cooling capabilities of CAF may have lead to firefighter injuries during interior fire attacks. Other firefighters thought that the bubble structure would provide the same mass flowrate of water with more surface area, which would enable a faster heat transfer rate and therefore, improve the cooling of the hot gas layer. There were also thoughts that the CAFS would coat the ceiling and eliminate energy feedback from the ceiling to the hot gas layer, thereby providing enhanced cooling.
